---
title: "Assignment 2 Cleaning"
format: html
editor: visual
author: Joshua Mull
date: 04/16/2025
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false  

# Libraries
library(tidyverse)

# file names
datadir_raw <- file.path("data","raw")

datadir_processed <- file.path("data", "processed/")

species_file <- "ASDN_Daily_species.csv"

snowsurvey_file <- "ASDN_Snow_survey.csv"
```

### Data Exploration

Import the snow survey

```{r}
# Import the species
snowsurvey_csv <- read_csv(file.path(datadir_raw, snowsurvey_file))

glimpse(snowsurvey_csv)

```

Character is where we would expect integer. Process to convert the Snow, Water and Land to the correct type

```{r}
# Check the unique values in the Snow_cover column
snowsurvey_csv %>%
  count(Snow_cover)
```

```{r}
# Identify non-numeric entries in Snow_cover column by attempting numeric conversion and filtering for NA
snowsurvey_csv %>% 
  count(Snow_cover) %>%
  filter(is.na(as.numeric(Snow_cover)))

# Make a copy and deal with the special characters
snowsurvey_fixed <- snowsurvey_csv

snowsurvey_fixed[snowsurvey_fixed == c(".", "-", "n/a", 'unk')] <- NA

# Make <1 = 0 instead of NA
snowsurvey_fixed[snowsurvey_fixed == "<1"] <- "0"
```

Now test to see if we only have NA's in the column 
```{r}
# Recheck for non-numeric entries in Snow_cover after replacements
snowsurvey_fixed %>%
  count(Snow_cover) %>%
  filter(is.na(as.numeric(Snow_cover)))
```

Now transform the Snow_cover column to numeric 
```{r}
# Convert the snow cover column to numeric 
snowsurvey_fixed <- snowsurvey_csv %>% 
  mutate(Snow_cover = as.numeric(Snow_cover))
```

Do the same for land and water 
```{r}
# Check the values for both 
snowsurvey_fixed %>%
  count(Water_cover)

snowsurvey_fixed %>%
  count(Land_cover)

```

Look at the non numeric
```{r}
# see which values are non numeric for the water and land cover columns 
snowsurvey_fixed %>% 
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))

snowsurvey_fixed %>% 
  count(Land_cover) %>%
  filter(is.na(as.numeric(Land_cover)))
```

Transform each the columns to numeric
```{r}
# transform each column to numeric just like the snow columnn
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = as.numeric(Land_cover))

snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = as.numeric(Water_cover))

glimpse(snowsurvey_fixed)
```

Check to make sure percent falls between 0 and 100
```{r}
# Check for values that are greater than 100% or are negative %
snowsurvey_fixed %>% 
  filter(Water_cover > 100 | Water_cover < 0) %>% 
  View()

snowsurvey_fixed %>% 
  filter(Land_cover > 100 | Land_cover < 0) %>% 
  View()
```

Set values above 100 and below 0 to NA

```{r}
# Set those values to NA because the cannot be true 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Land_cover > 100, NA, Land_cover))

snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Land_cover = ifelse(Water_cover > 100, NA, Water_cover))

# also include the snow column from lecture 
snowsurvey_fixed <- snowsurvey_fixed %>% 
  mutate(Snow_cover = ifelse(Snow_cover > 100, NA, Snow_cover))
```

Write to a csv file 
```{r}
# check if the folder exists
dir.create(datadir_processed, showWarnings = FALSE)

# write the file
write_csv(snowsurvey_fixed, file.path(datadir_processed, "all_cover_fixed_Mull.csv"))
```

